Young protoplanetary disks

### 2012-12-03 paper.tex
3/12/12

\title[Simulated observations of young protoplanetary disks]{Young protoplanetary disks}
\author[Tom  Douglas, Paola Caselli, Et al.]{Tom Douglas$^{1}$\thanks{E-mail:
pytd@leeds.ac.uk}, Paola Caselli$^{1}$\\
$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK}
\begin{document}

@@

\subsection{Grid Construction} \label{subsec:gridding}
In order to construct the grid, points are randomly selected from the volume, then compared against a reference point. Grid points are selected at random in cylindrical co-ordinates, linearly spaced in z and $\phi$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random co-ordinates, the hydrogen density and molecular density at the point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use, if not then another r, $\phi$, z co-ordinate is selected. The weighting function and point distribution were selected empirically to sample all the scales while ensuring that the majority of points are located in the inner disc where the density is higher. 20\% of these points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radius of the model) in order to stop too many of the selecected points clustering in the high density disc and leaving the envelope undersampled. In addiation to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See figure \ref{points} for an example of the points distribution in r, z. \newline

@@

NOTE: YOU NEED TO ADD A LITTLE HISTORY ON WHY YOU SELECTED THESE SPECIES TO DO THE RADIATIVE TRANSFER. CAN YOU PLEASE RESTRUCTURE THIS SECTION ?
%Figures showing the RT results in a few molecules/transitions (CO, HCO$^+$, HCN, OCS, H2CO, NH3 --- maybe H2O,H2$^{18}$O [to try first])
The results of these simulations for the molecules OCS, HCO$^+$, H$_2$CO and isotopologues of CO are shown here, other molecules simulated but not shown include HCN, HNC, HNO, HCS$^+$ and  CS.
For the purpose of simulating observations the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc) and is inclined 30$^\circ$ to edge on. From these observations integrated intensity, intensity weighted velocity and position velocity diagrams through the centre of the model were created.
The simulations done are focused upon frequencies within ALMA band 7, selected to give the best trade off between resolution and sensitivity for early ALMA science.
(Note the moment 1 and 0 maps were created by integrating between -12.5 to -0.5 km$\,$s$^{-1}$ and +0.5 to +12.5 km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope, this can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity, moment 1 maps are shown with a cutoff of 1/1000 of the peak emission/absorption value)\newline

@@

Figures 6,7 and 8 show synthetic images of the CO J=3-2 line at 345.8$\,$GHz. As the upper level for this line is at 33$\,$K above the ground state this transition should be excited through out disc. given the large and uniform nature of the abundance of CO in the disc this line will be optically thick and this is only mapping the outer regions of the disc. In common with most transitions simulated, the line is seen almost exclusively in absorption against the disc continuum. The region of emission in the centre is where there is no continuum background to be absorbed.  Some indications of spiral structure can be seen but are indistinct. ***not sure how useful these are, might be worth cutting them?***\newline

@@

After looking at a variety of lines which could be observed of ALMA band 6 or 7 (211-275$\,$GHz and 275-373$\,$GHz) we found that the OCS lines in alma band 7 (22-21 through 30-29) with upper energy levels between 161 and 271 K above the ground state provide a way to trace hot shocked gas in spiral arms without resolving structure. Figures \ref{OCS_mom0} \& \ref{smoothOCS_mom0} show the integrated intensity of the OCS J=28-27 line in the model described previously and one where the disc section of the model is an axis-symmetric model created by taking the azimuthal average of density, temperature and molecular abundance. The order of magnitude difference in the line intensity provides a method to detect hot dense gas without being able to resolve the spiral structure spatially.\newline

\begin{figure}
 \includegraphics[width=84mm]{Figures/sim/imagesmoothedOCS_28-27_30deg_contSub.png}
 \label{smoothOCS_mom0}
 \caption{SmoothedOCS 28-27 Continuum subtracted mom0}
\end{figure}

@@

More complex molecules such as H$_2$CO with many closely spaced spectral lines can be used to gain an estimate on the temperature of a region.

(figure of temperature as a function of line ratio for 2 sets of formaldehyde lines and map of derived temperature, talk to P about this. This doesn't seem doable with the lines around 1mm for alma)\newline

@@

Some molecules such as HCO$^+$ trace only the outer regions of the disc (Ilee 2011) and so can be used to look at the extended velocity and physical structure. In these colder less dense regions we see the molecular lines in emission rather than absorption. Figure \ref{hcop_pv} shows the rotation curve of the model as shown in figure \ref{velocity} against the HCO$^+$ J=3-2 line emission. It is clear that from observations such as these the rotation curve of a disc could be reconstructed.\newline
\newline

In all the simulations except HCO$^+$ we see molecular lines in absorption throughout the majority of the disc and in some cases a small amount of emission towards the centre, where the inclination of the disc allows us the view the low density but moderate temperature region inside the inner edge of the spiral structures without any high temperature and density material behind it.\newline

@@

In this paper we have performed radiative transfer simulations of a hybrid model comprising a 0.4$\, M_\odot$ self gravitating disc with radius 64$\,AU$ showing spiral density waves, surrounded by an envelope simulated as a collapsing 10$\,M_\odot$ BE-sphere. CASA simulations show that at a distance of 100$\,$pc both spatial resolution of the spirals in such a disc and extraction of kinematic and temperature from molecular lines are possible in ALMA band 7. Our simulations show that many  molecular species are predominantly seen in absorption towards the centre of  self-gravitating protoplanetary discs. The quiescent nature of the envelope around such discs only obscures them within $\pm\,0.5 km\,s^{-1}$.\newline
Hot, shocked gas can be detected in such a disc without resolving it by observing OCS transitions between 26-27 to 30-29.\newline
One assumption made in this model is that the gas and dust are in thermal equilibrium.If they are not and the dust is significantly cooler than the gas then transitions may not show up in absorption.\newline
The mass of the disc used (0.4$\,\rm{M}_\odot$) is on the high end for early discs (citations for estimate of early class 1 disc masses).\newline 
In the model spirals shock heat the mid-plane layer and the colder more diffuse gas absorbs the continuum from it, this results in molecular lines being seen in absorbtion whenever lines of sight pass through the cold diffuse gas at larger heights on to the hot dense midplane of the disc. Rotation curves can be gathered from these observations even with spiral structure\newline

@@

%
%(This appendix was not part of the original paper by
%A.V.~Raveendran and is included here just for illustrative
%purposes. The references are not relevant to the text of the
%appendix, they are references from the bibliography used to
%illustrate text before and after citations.)
%
%Spectroscopic observations of bright quasars show that the mean
%number density of Ly$\alpha$ forest lines, which satisfy certain
%criteria, evolves like $\rmn{d}N/\rmn{d}z=A(1+z)^\gamma$, where
%$A$ and~$\gamma$ are two constants.  Given the above intrinsic
%line distribution we examine the probability of finding large gaps
%in the Ly$\alpha$ forests.  We concentrate here only on the
%statistics and neglect all observational complications such as the
%line blending effect \citep[see][for example]{b11}.
%
%Suppose we have observed a Ly$\alpha$ forest between redshifts $z_1$
%and~$z_2$ and found $N-1$ lines.  For high-redshift quasars $z_2$~is
%usually the emission redshift $z_{\rmn{em}}$ and $z_1$ is set to
%$(\lambda_{\rmn{Ly}\beta}/\lambda_{\rmn{Ly}\alpha})(1+z_{\rmn{em}})=0.844
%(1+z_{\rmn{em}})$ to avoid contamination by Ly$\beta$ lines.  We
%want to know whether the largest gaps observed in the forest are
%significantly inconsistent with the above line distribution.  To do
%this we introduce a new variable~$x$:
%%
%\begin{equation}
%x={(1+z)^{\gamma+1}-(1+z_1)^{\gamma+1} \over
%     (1+z_2)^{\gamma+1}-(1+z_1)^{\gamma+1}}.
%\end{equation}
%%
%$x$ varies from 0 to 1.  We then have $\rmn{d}N/\rmn{d}x=\lambda$, where $\lambda$
%is the mean number of lines between $z_1$ and $z_2$ and is given by
%%
%\begin{equation}
%\lambda\equiv{A[(1+z_2)^{\gamma+1}-(1+z_1)^{\gamma+1}]\over\gamma+1}.
%\end{equation}
%%
%This means that the Ly$\alpha$ forest lines are uniformly
%distributed in~$x$. The probability of finding $N-1$ lines between $z_1$
%and~$z_2$, $P_{N-1}$, is assumed to be the Poisson distribution.
%%
%\newpage
%%
%\begin{figure}
%\vspace{11pc}
%\caption{$P(>x_{\rmn{gap}})$ as a function of $x_{\rmn{gap}}$ for,
% from left to right, $N=160$, 150, 140, 110, 100, 90, 50, 45 and~40.
% Compare this with \protect\citet{b15}.}
%\label{appenfig}
%\end{figure}
%
%\subsection{Subsection title}
%
%We plot in Fig.~\ref{appenfig} $P(>x_{\rmn{gap}})$ for several $N$
%values. We see that, for $N=100$ and $x_{\rmn{gap}}=0.06$,
%$P(>0.06)\approx20$ per cent.  This means that the probability of
%finding a gap with a size larger than six times the mean
%separation is not significantly small. When the mean number of
%lines is large, $\lambda\sim N>>1$, our $P(>x_{\rmn{gap}})$
%approaches the result obtained by \citet[fig. 4]{b22} for small
%(but still very large if measured in units of the mean separation)
%$x_{\rmn{gap}}$, i.e., $P(>x_{\rmn{gap}})\sim N(1-
%x_{\rmn{gap}})^{N-1}\sim N {\rmn{exp}}(-\lambda x_{\rmn{gap}})$.
%
\bsp
%
\label{lastpage}
%
\end{document}

>>>>

\title[Simulated observations of young protoplanetary disks]{Young protoplanetary disks}
\author[Tom  Douglas, Paola Caselli, Et al.]{Tom Douglas$^{1}$\thanks{E-mail:
pytd@leeds.ac.uk}, Paola Caselli$^{1}$, Et al.\\
$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK}
\begin{document}

@@

\subsection{Grid Construction} \label{subsec:gridding}
In order to construct the grid, points are randomly selected from the volume then have the density and molecular density at the point compared against a reference point in order to decide if the point is to be accepted or not. Grid points are selected at random in cylindrical co-ordinates, linearly spaced in z and $\phi$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) as a threshold. After selection of random co-ordinates, the hydrogen density and molecular density at the point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use, if not then another r, $\phi$, z co-ordinate is selected. The weighting function and point distribution were selected empirically to sample all the scales while ensuring that the majority of points are located in the inner disc where the density is higher. 20\% of these points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radius of the model) in order to stop too many of the selecected points clustering in the high density disc and leaving the envelope undersampled. In addiation to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See figure \ref{points} for an example of the points distribution in r, z. \newline

@@

Simulated molecules were confined to speices where the chemical abundance in the disc and envelope are known, or have simulated abundances. In addition radative transfer simulations are limited to are limited to molecules where the collisional and Einstein co-efficients are known. Lines for these molecules were simulated with LIME for freqencies 50$\sim$500$\,$GHz, with focus on lines which could be observed in ALMA band 7. The results presented in this section represent those lines which show dectectable emission/absorption which can be used to trace either spiral structure or rotation.\newline

%Figures showing the RT results in a few molecules/transitions (CO, HCO$^+$, HCN, OCS, H2CO, NH3 --- maybe H2O,H2$^{18}$O [to try first])
The results of these simulations for the molecules OCS, HCO$^+$, H$_2$CO and isotopologues of CO are shown here, other molecules simulated but not shown include HCN, HNC, HNO, HCS$^+$ and  CS.
For the purpose of simulating observations the model was placed at roughly the distance of nearby low-mass star forming regions (100$\,$pc) and is inclined at 30$^\circ$ to edge on. From these observations integrated intensity, intensity weighted velocity and position velocity diagrams through the centre of the model were created.
The simulations done are focused upon frequencies within ALMA band 7, selected to give the best trade off between resolution and sensitivity for early ALMA science {\bf IS BAND 7 STILL THE BEST FOR US TO LOOK AT SEEING AS WE ARE NOW LOOKING AT ALMA FULL CAPABILITY RATHER THAN CYCLE1?}.
(Note the moment 1 and 0 maps were created by integrating between -12.5 to -0.5 km$\,$s$^{-1}$ and +0.5 to +12.5 km$\,$s$^{-1}$ to avoid being dominated by the contribution from the envelope, this can be seen in some PV diagrams as the strong absorption feature at all positions around zero velocity, moment 1 maps are shown with a cutoff of 1/1000 of the peak emission/absorption value)\newline

@@

Figures 6,7 and 8 show synthetic images of the CO J=3-2 line at 345.8$\,$GHz. As the upper level for this line is at 33$\,$K above the ground state this transition should be excited through out disc. given the large and uniform nature of the abundance of CO in the disc this line will be optically thick and this is only mapping the outer regions of the disc. In common with most transitions simulated, the line is seen almost exclusively in absorption against the disc continuum. The region of emission in the centre is where there is no continuum background to be absorbed.  Some indications of spiral structure can be seen but are indistinct. {\bf not sure how useful these are, might be worth cutting them?}\newline

@@

After looking at a variety of lines which could be observed of ALMA band 6 or 7 (211-275$\,$GHz and 275-373$\,$GHz) we found that the OCS lines in alma band 7 (22-21 through 30-29) with upper energy levels between 161 and 271 K above the ground state provide a way to trace hot shocked gas in spiral arms without resolving structure. Figures \ref{OCS_mom0} \& \ref{smoothOCS_mom0} show the integrated intensity of the OCS J=28-27 line in the model described previously and one where the disc section of the model is an axis-symmetric model created by taking the azimuthal average of density, temperature and molecular abundance. The order of magnitude difference in the line intensity provides a method to detect hot dense gas without being able to resolve the spiral structure spatially.\newline
%
%\begin{figure}
% \includegraphics[width=84mm]{Figures/sim/imagesmoothedOCS_28-27_30deg_contSub.png}
% \label{smoothOCS_mom0}
% \caption{SmoothedOCS 28-27 Continuum subtracted mom0}
%\end{figure}

@@

More complex molecules such as H$_2$CO with many closely spaced spectral lines can be used to gain an estimate on the temperature of a region.\newline

@@

Some molecules such as HCO$^+$ trace only the outer regions of the disc (Ilee 2011) and so can be used to look at the extended velocity and physical structure. In these colder less dense regions we see the molecular lines in emission rather than absorption. Figure \ref{hcop_pv} shows the rotation curve of the model as shown in figure \ref{velocity} against the HCO$^+$ J=1-0 line emission. It is clear that from observations such as these the rotation curve of a disc could be reconstructed.\newline

In all the simulations except HCO$^+$ we see molecular lines in absorption throughout the majority of the disc and in some cases a small amount of emission towards the centre. Molecular lines throughout the plane of the disc show up in absorption against the continuum emission of the shock heated, dense midplane.

@@

In this paper we have performed radiative transfer simulations of a hybrid model comprising a 0.4$\, M_\odot$ self gravitating disc with radius 64$\,AU$ showing spiral density waves, surrounded by an envelope simulated as a collapsing 10$\,M_\odot$ BE-sphere. CASA simulations show that at a distance of 100$\,$pc both spatial resolution of the spirals in such a disc and extraction of kinematic and temperature from molecular lines are possible in ALMA band 7. Our simulations show that many  molecular species are predominantly seen in absorption towards the centre of  self-gravitating protoplanetary discs. The quiescent nature of the envelope around such discs only obscures them within $\pm\,0.5 km\,s^{-1}$.\newline
One assumption made in this model is that the gas and dust are in thermal equilibrium.If they are not and the dust is significantly cooler than the gas then transitions may not show up in absorption.\newline
%The mass of the disc used (0.4$\,\rm{M}_\odot$) is on the high end for early discs (citations for estimate of early class 1 disc masses).\newline
In the model spirals shock heat the mid-plane layer and the colder more diffuse gas absorbs the continuum from it, this results in molecular lines being seen in absorbtion whenever lines of sight pass through the cold diffuse gas at larger heights on to the hot dense midplane of the disc. Rotation curves can be gathered from these observations even with spiral structure\newline
Outflows could contaminate measurements of rotation curves meaning that speices  which commonly trace outflows, such as C$^{16}$O and HC$^{16}$O$^+$ are not going to be good tracers of disc rotation. However speices such as OCS and C$^{18}$O which are not seen in outflows can be used to trace disc rotation.

@@

\bsp
%
\label{lastpage}
%
\end{document}
